---
title: Sorting Comparison
author: John Lee
date: '2019-03-08'
slug: sorting-comparison
categories:
  - coding
tags:
  - coding
subtitle: ''
---
As I'm self studying algorithms and data structures with python from [here](https://runestone.academy/runestone/static/pythonds/index.html), I figured I could try to do some experiments with different sorting algorithms using my own implementations in R.

Types of sorting algorithms I will use:

- Bubble Sort
- Insertion Sort
- Selection Sort
- Shell Sort
- Merge Sort
- Quick Sort

I will be dealing with a vector of type `double`. It can be a collection of any real positive numbers.
```{r seed}
set.seed(3706)
```

The following function is used as a sanity check to confirm that each function sorts the given vector as it should.
```{r order_check_function}
check_works_correct <- function(myFunc){
  tmp <- function(myfunc) {
    x <- sample(100, 50) # integers; This checks that the algorithm works as it should even with duplicates
    y <- runif(50, 0, 500)
    return(identical(sort(x), myfunc(x)) && identical(sort(y), myfunc(y))) # compare to R's sort function
  }
  return(all(replicate(n = 100, tmp(myFunc))))
}
```

# Write Sorting Algorithms

**Bubble Sort**
```{r Bubble_Sort}
bubble_sort <- function(x) {
  pass_remain <- length(x)
  swapped <- TRUE
  while ((pass_remain >= 1) & swapped) {
    swapped <- FALSE
    a <- 1
    b <- 2
    while (a < pass_remain) {
      if (x[a] > x[b]) {
        tmp <- x[a]
        x[a] <- x[b]
        x[b] <- tmp
        swapped <- TRUE
      }
      a <- a + 1
      b <- b + 1
    }
    pass_remain <- pass_remain - 1
  }
  return(x)
}

check_works_correct(bubble_sort)
```

**Insertion Sort**
```{r insertion_sort}
insertion_sort <- function(x) {
  for (i in 2:length(x)) {
    item <- x[i]
    pos <- i
    while ((x[pos - 1] > item) && (pos > 1)) {
      x[pos] <- x[pos - 1]
      pos <- pos - 1
    }
    x[pos] <- item
  }
  return(x)
}

check_works_correct(insertion_sort)
```

**Selection Sort**
```{r selection_sort}
selection_sort <- function(x) {
  for (i in seq(length(x), 1, -1)) {
    max_ind <- 1
    for (j in 1:i) {
      if (x[j] > x[max_ind]) {
        max_ind <- j
      }
    }
    tmp <- x[i]
    x[i] <- x[max_ind]
    x[max_ind] <- tmp
  }
  return(x)
}

check_works_correct(selection_sort)
```

**Shell Sort**
```{r shell_sort}
shell_sort <- function(x) {
  # choose max increment (Hibbard Increment)
  k <- floor(log(length(x)+1, 2))
  # sort all the sublists with the increment, while decreasing the increment
  while (k > 0) {
    gap <- 2^k - 1
    # sort each sublist with insertion sort
    for (i in 1:gap) {
      compare_pos <- i + gap
      while (compare_pos < (length(x) + 1)) {
        item <- x[compare_pos]
        pos <- compare_pos
        while (x[pos-gap] > item && pos > i) {
          x[pos] <- x[pos-gap]
          pos <- pos-gap
        }
        x[pos] <- item
        compare_pos <- compare_pos + gap
      }
    }
    # decrement
    k <- k - 1
  }
  return(x)
}

check_works_correct(shell_sort)
```

**Merge Sort**
```{r merge_sort}
# merge sort with recursion
merge_sort <- function(x) {
  if (length(x) < 2) { # base case
    return(x)
  } else {
    break_point <- length(x) %/% 2
    
    # break the vector into two
    left <- x[1:break_point]
    right <- x[(break_point+1):length(x)]
    
    # recurecurecursionsionsion
    left <- merge_sort(left)
    right <- merge_sort(right)
    
    insert_pos <- 1
    left_pos <- 1
    right_pos <- 1
    
    # merge left and right
    while (left_pos <= length(left) && right_pos <= length(right)) {
      if (left[left_pos] <= right[right_pos]) {
        x[insert_pos] <- left[left_pos]
        left_pos <- left_pos + 1
      } else {
        x[insert_pos] <- right[right_pos]
        right_pos <- right_pos + 1
      }
      insert_pos <- insert_pos + 1
    }
    
    if (left_pos <= length(left)) {
      while (left_pos <= length(left)) {
        x[insert_pos] <- left[left_pos]
        left_pos <- left_pos + 1
        insert_pos <- insert_pos + 1
      }
    }
    
    if (right_pos <= length(right)) {
      while (right_pos <= length(right)) {
        x[insert_pos] <- right[right_pos]
        right_pos <- right_pos + 1
        insert_pos <- insert_pos + 1
      }
    }
    
    return(x)
  }
}
check_works_correct(merge_sort)
```

**Quick Sort**
```{r quick_sort}
quick_sort <- function(x) {
  low <- 1
  high <- length(x)
  x <- quick_sort_helper(x, low, high)
  return(x)
}

quick_sort_helper <- function(x, low, high) {
  if (low >= high) {
    return(x)
  } else {
    # choose pivot
    mid <- (low + high) %/% 2
    if ((x[low] < x[mid] && x[mid] < x[high]) || (x[high] < x[mid] && x[mid] < x[low])) {
      median_ind <- mid
    } else if ((x[mid] < x[low] && x[low] < x[high]) || (x[high] < x[low] && x[low] < x[mid])) {
      median_ind <- low
    } else {
      median_ind <- high
    }
    
    # put pivot to last index
    pivot <- x[median_ind]
    if (median_ind != high) {
      x[median_ind] <- x[high]
      x[high] <- pivot
    } 
    
    left <- low
    right <- high-1
    
    # rearrange the values around the pivot
    done <- FALSE
    while (!done) {
      while (left <= right && x[left] < pivot) {
        left <- left + 1
      }
      while (right >= left && x[right] >= pivot) {
        right <- right - 1
      }
      if (left > right) {
        done <- TRUE
      } else {
        tmp <- x[right]
        x[right] <- x[left]
        x[left] <- tmp
      }
    }
    x[high] <- x[left]
    x[left] <- pivot
    
    x <- quick_sort_helper(x, low, left-1)
    x <- quick_sort_helper(x, left+1, high)
  }
  return(x)
}

check_works_correct(quick_sort)
```

Now I'll compare the runtime of each sort, varying the size and shape of the graphs.

# Generate Unordered Sequences with Varying Trends

I built the function `generate_random_sequence`, which accepts the following to generate a random sequence:

- `trend`: a function that describes the general trend of the sequence of the sequences
- `n`: number of points
- `low`: lowest `x` value. Defaults to 0.
- `high`: highest `x` value. Defaults to 10.
- `noise`: a function to be used to generate noise for the sequence. Defaults to a Gaussian Distribution with mean 0 and sd 5

```{r random_sequence_generator}
generate_random_sequence <- function(trend, n, low=0, high=10, noise = function(x) rnorm(x, 0, 5)) {
  # trend: a function; general trend of the sequence
  # noise: a function; amount of noise you want to add to a sequence
  x <- trend(seq(low, high, length.out = n)) + noise(n)
  if (any(x <= 0)) {
    x <- x + abs(min(x)) + 1
  }
  return(x)
}
```

For example, suppose I want to generate a sequence of length `n = 100` that follows a quadratic trend. (I put less emphasis on the aesthetics of the graphs. I just wanted to quickly demonstrate the graphs.)
```{r example1}
trend1 <- function(x) (x-5)^2 # quadratic trend
x1 <- generate_random_sequence(trend1, 100)

par(mfrow = c(1, 2), mar = c(1, 0.1, 2, 2), oma=c(1.5,2,2,1))
plot(x1, xaxt = 'n', main = 'Scatterplot', xlab='')
barplot(x1, main = 'Barplot')
mtext(expression(bold("Sequence with Quadratic Trend")), outer=TRUE, cex=1.5)
```

Create various types of trends
```{r create_trends}
quadratic_trend <- function(x) (x-5)^2
right_skew_trend <- function(x) dgamma(x, 4, 2) * 10
left_skew_trend <- function(x) dgamma(10-x, 4, 2) * 10
symmetrical_trend <- function(x) dnorm(x, 5, 1.3) * 10
flat_trend <- function(x) dunif(x, 0, 10) * 10
sine_trend <- function(x) sin(x*2) * 5 

trend_list <- list(
  # all range from 0 to 10
  quadratic_trend = quadratic_trend,
  right_skew_trend = right_skew_trend, 
  left_skew_trend = left_skew_trend, 
  symmetrical_trend = symmetrical_trend, 
  flat_trend = flat_trend, 
  sine_trend = sine_trend
)

par(mfrow = c(1, 2))
for (i in 1:length(trend_list)) {
  plot(trend_list[[i]], 0, 10, main = names(trend_list)[i], xaxt = "n", xlab = "")
}
```

What they look like when noise is added:
```{r barplots, eval=FALSE}
par(mfrow = c(1,2))
barplot(generate_random_sequence(quadratic_trend, 100, noise = function(x) rnorm(x, 0, 5)), main = "Quadratic Trend")
barplot(generate_random_sequence(right_skew_trend, 100, noise = function(x) rnorm(x, 0, 1)), main = "Right Skew Trend")
barplot(generate_random_sequence(left_skew_trend, 100, noise = function(x) rnorm(x, 0, 1)), main = "Left Skew Trend")
barplot(generate_random_sequence(symmetrical_trend, 100, noise = function(x) rnorm(x, 0, 0.7)), main = "Symmetrical Trend")
barplot(generate_random_sequence(flat_trend, 100, noise = function(x) rnorm(x, 0, 0.8)), main = "Flat Trend")
barplot(generate_random_sequence(sine_trend, 100, noise = function(x) rnorm(x, 0, 2)), main = "Sine Trend")
```
```{r barplots_evalled, echo=FALSE}
par(mfrow = c(1,2))
barplot(generate_random_sequence(quadratic_trend, 100, noise = function(x) rnorm(x, 0, 5)), main = "Quadratic Trend")
barplot(generate_random_sequence(right_skew_trend, 100, noise = function(x) rnorm(x, 0, 1)), main = "Right Skew Trend")
barplot(generate_random_sequence(left_skew_trend, 100, noise = function(x) rnorm(x, 0, 1)), main = "Left Skew Trend")
barplot(generate_random_sequence(symmetrical_trend, 100, noise = function(x) rnorm(x, 0, 0.7)), main = "Symmetrical Trend")
barplot(generate_random_sequence(flat_trend, 100, noise = function(x) rnorm(x, 0, 0.8)), main = "Flat Trend")
barplot(generate_random_sequence(sine_trend, 100, noise = function(x) rnorm(x, 0, 2)), main = "Sine Trend")
```

Also, consider some special cases, such as already sorted, sorted backwards, etc.
```{r construct_special_cases}
already_sorted <- function(n) seq(1, n)
sorted_backwards <- function(n) seq(n, 1, -1)
one_off <- function(n) {
  mySeq <- seq(1, n)
  random_index <- sample(mySeq, 2)
  mySeq[random_index] <- mySeq[random_index[c(2,1)]]
  return(mySeq)
}
extreme_out_of_place <- function(n, lowest=TRUE) {
  mySeq <- seq(1,n)
  index <- ifelse(lowest, 1, length(mySeq))
  replace_val <- ifelse(lowest, n+1, 0)
  mySeq[index] <- replace_val
  return(mySeq)
}

special_case_list <- list(
  already_sorted = already_sorted,
  sorted_backwards = sorted_backwards,
  one_off = one_off,
  extreme_out_of_place_low = function(x) extreme_out_of_place(x, TRUE),
  extreme_out_of_place_high = function(x) extreme_out_of_place(x, FALSE)
)

par(mfrow = c(1,2))
for (i in 1:length(special_case_list)) {
  barplot(special_case_list[[i]](100), main = names(special_case_list)[i], cex.main = 0.9)
}
```

# Sorting with Different Algorithms

Now that I have the tools to construct random sequences, I can finally start generating sequences and run benchmarks.

For future convenience, I saved the results in two different lists: `trend_sort_time` and `special_case_sort_time`. Each list are divided by number of unsorted sequences (10, 100, 1000). And for each number are microbenchmarks for each "trends" or "special cases".

`microbenchmark` is used to precisely evaluate the performance of each algorithm. Each sorting algorithm sorts randomly generated data 100 times. On each iteration, sequences are randomly generated, and each algorithm sorts the same sequence.

Microbenchmark for trends:

```{r trend_microbenchmark, cache=TRUE}
library(microbenchmark)

# each trend uses different noise scale.
noise_list <- list(
  "quadratic_trend" = function(x) rnorm(x, 0, 5),
  "right_skew_trend" = function(x) rnorm(x, 0, 1),
  "left_skew_trend" = function(x) rnorm(x, 0, 1),
  "symmetrical_trend" = function(x) rnorm(x, 0, 0.7),
  "flat_trend" = function(x) rnorm(x, 0, 0.8),
  "sine_trend" = function(x) rnorm(x, 0, 2)
)

N <- c(10, 100, 1000) # size of sequence

# loop through each N and trends
trend_sort_time <- setNames(
  rep(list(setNames(vector("list", length(trend_list)), names(trend_list))), length(N)),
  N)
print(str(trend_sort_time))

for (n in N) {
  for (trend in names(trend_list)) {
    trend_sort_time[[as.character(n)]][[trend]] <-
      microbenchmark(
        "bubble_sort" = bubble_sort(Seq),
        "insertion_sort" = insertion_sort(Seq),
        "selection_sort" = selection_sort(Seq),
        "shell_sort" = shell_sort(Seq),
        "merge_sort" = merge_sort(Seq),
        "quick_sort" = quick_sort(Seq),
        times = 100L,
        setup = assign("Seq", 
                       generate_random_sequence(
                         trend=trend_list[[trend]],
                         n=n,noise = noise_list[[trend]]))
      )
  }
}

print(trend_sort_time)
```

Microbenchmark for Special Cases
```{r special_case_benchmark, cache=TRUE}
special_case_sort_time <- setNames(
  rep(list(setNames(vector("list", length(special_case_list)), names(special_case_list))), length(N)),
  N)
print(str(special_case_sort_time))

for (n in N) {
  for (sc in names(special_case_list)) {
    special_case_sort_time[[as.character(n)]][[sc]] <-
      microbenchmark(
        "bubble_sort" = bubble_sort(Seq),
        "insertion_sort" = insertion_sort(Seq),
        "selection_sort" = selection_sort(Seq),
        "shell_sort" = shell_sort(Seq),
        "merge_sort" = merge_sort(Seq),
        "quick_sort" = quick_sort(Seq),
        times = 100L,
        setup = assign("Seq", special_case_list[[sc]](n))
      )
  }
}

print(special_case_sort_time)
```
```{r, echo=FALSE}
save(trend_sort_time, special_case_sort_time, file="sort_comparisons")
```

It's not easy to detect any obvious trend with only numbers, so in the next post, I'll try to find ways to visualize these results to make differences more apparent.